Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study

### 2025-07-02 inst/paper/main.tex
add supplementary methods

>>>>

\subsection*{Supplementary Methods}

\subsubsection*{Pandemic potential influenza subtypes}

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}. H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}. The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}. Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}. For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes. This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.

\subsubsection*{Transmission model generation time from incubation period}

Here we provide a more detailed description of how the generation time is calculated from the incubation period. The time between an infector being infected and then infecting an infectee (i.e. generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020, lintonCorrelationTimesSARSCoV22022}. This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived. This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}. \\

Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C). The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution). In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector. \\

We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: 1\%, 15\%, 30\%, these correspond to $\alpha$ values of 31.82053, 1.962614, and 0.726558, respectively (Figure \ref{fig:delay-distributions}C).

\subsubsection*{Modelling non-pharmaceutical interventions}

The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility). Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission. The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e. general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}. Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}). When a case is symptomatic they are isolated after a delay. The timing of isolation is drawn from the onset-to-isolation distribution. Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times. In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{iso}$ is zero. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{iso}$). Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated. The isolation of the infectee is whatever occurs first of these two events.

\subsubsection*{Defining outbreak control}

Outbreak control, or the extinction or elimination of the pathogen, is a binary classification, which is true if there are no new cases from the 12th week of the outbreak onwards, and false otherwise. This is the same definition as used in \cite{hellewellFeasibilityControllingCOVID192020}. The 12 week threshold is somewhat arbitrary but is judged to be long enough that the NPIs can bring the effective reproduction number below one for all transmission to cease. The number of generations in an outbreak lasting 12 weeks will be approximately between 18 generations (H5N1 median incubation period and 0.1\% presymptomatic transmission) and 84 (H1N1 median incubation period and 30\% presymptomatic transmission), although realisations of the stochastic simulation may be greater or less than these estimates from distribution medians. \\

The maximum number of cases in each simulation replicate before terminating the simulation and concluding it is uncontrolled was set to 5,000. This number should be larger than the expected cluster size from a subcritical outbreak which is given by $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number. For example, using the largest initial outbreak size in this study $N_i = 40$, and assuming $R$ is subcritical but very close to unity, $R = 0.99$, gives a possible outbreak size of 4,000 cases. Therefore, our limit that 5,000 cases is an uncontrolled outbreak should not include any outbreaks that will go to extinction eventually but produce transient outbreaks by chance. \\

For computational efficiency we capped the maximum number of days the simulation could run for to 140 (20 weeks). This is substantially longer than the 12 cutoff evaluating outbreak control so does not influence the results.

\subsection*{Supplementary Results}
